---
title: "<b>Developing Data Products: Final Pitch</b>"
subtitle: "Use the arrow keys to navigate slides"
author: "Chris McKelt"
date: "October 2017"
output:
  revealjs::revealjs_presentation:
    theme: solarized
    highlight: pygments
    center: true
---

```{r setup, include=FALSE}
suppressMessages(rm(list = ls()))
suppressMessages(setwd("C:/dev/r-course/9-data-products/pitch"))
source("c:/dev/r-course/include.r")
using("RSQLite")
using("sqldf")
using("choroplethr")
using("choroplethrMaps")
using("DescTools")
using("readxl")
using("devtools")
using("ggplot2")
using("plotly")
using("tm")
using("RColorBrewer")
using("wordcloud")


Sys.setenv("plotly_username" = "chrismckelt")
Sys.setenv("plotly_api_key" = "M6S961nyr6MaEAwtNAM0")
options(DT.fillContainer = FALSE)
options(DT.autoHideNavigation = FALSE)
options(scipen = 999)
set.seed(3333)

```

## What is this application?

### Stage 1 
Analysis over the Lending Clubs loan book data.  It allows the contrast of a dataset variable against the aggregated loan amount.

### Stage 2 (to be done post investment)
Application will take a loan application and predict its probability of default and a maximum loanable amount according to the applicants credit grade.

## Overview of the Lending Club dataset

[Data source](https://www.kaggle.com/wendykan/lending-club-loan-data)

The data contains all loans issued through the 2007-2015, including the current loan status (Current, Late, Fully Paid, etc.) and latest payment information.

The file containing loan data through the "present" contains complete loan data for all loans issued through the previous completed calendar quarter. Additional features include credit scores, number of finance inquiries, address including zip codes, and state, and collections among others. 

The file is a matrix of about 890 thousand observations and 75 variables. A data dictionary is provided in a separate file.


## What loans are people applying for in the Lending Club dataset?

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
connection_string <- 'C:/dev/r-course/9-data-products/week-4/data/database.sqlite'
db <- RSQLite::dbConnect(SQLite(), dbname = connection_string, loadable.extensions = TRUE, cache_size = NULL, synchronous = "off", flags = SQLITE_RWC, vfs = NULL)

data.tables = dbListTables(db)
data_loanbook <- sqldf("select * from loan", connection = db)
data_loanbook$issue_d <- as.Date(gsub("^", "01-", data_loanbook$issue_d), format = "%d-%b-%Y")
data_loanbook$grade <- as.factor(data_loanbook$grade)

data_loanbook.corpus <- Corpus(DataframeSource(data.frame(head(data_loanbook[, 23], n = 10000))))
data_loanbook.corpus <- tm_map(data_loanbook.corpus, removePunctuation)
data_loanbook.corpus <- tm_map(data_loanbook.corpus, content_transformer(tolower))
```

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
wordcloud(data_loanbook.corpus,
          max.words = 100,
          random.order = FALSE,
          rot.per = 0.30,
          use.r.layout = FALSE,
          colors = brewer.pal(8, "Paired"))
```

## Where to from here?




## References

[Kaggle](https://www.kaggle.com/wendykan/lending-club-loan-data)</br>


###rmarkdown::render('C:/dev/r-course/9-data-products/pitch/pitch.Rmd')