---
title: "Statistical Inference Project"
author: Chris McKelt
output:
  pdf_document: default
  html_document: default
---

## Part 2: Basic Inferential Data Analysis

#### Description: https://www.coursera.org/learn/statistical-inference/peer/3k8j5/statistical-inference-course-project
#### https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html
#### ToothGrowth: The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. 
#### Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, 
#### orange juice or ascorbic acid (form  of vitamin C and coded as VC).





```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
  rm(list = ls()) # clear vars
  setwd("C:\\dev\\r-course\\5-statistical-inference")

  list.of.packages <- c("dplyr", "tidyr", "ggplot2", "knitr", "markdown", "moments", "nortest", "e1071", "data.table", "sqldf", "pastecs")
  new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
  
  if (length(new.packages))
      install.packages(new.packages)
  
  sapply(sapply(list.of.packages, library, character.only = TRUE, quietly = TRUE), require, character.only = TRUE, quietly = TRUE)

```


```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# utility function to test if pvalue > alpha & output whether to reject H0
  test_reject <- function(ho, tresult) {
      alpha <- .05
      if (tresult$p.value > alpha) {
          #If the p-value is less than or equal to the chosen significance level (?), 
          #the test suggests that the observed data is inconsistent with the null hypothesis, 
          #so the null hypothesis must be rejected. However, that does not prove that the tested hypothesis is true
          msg <- paste("p.value is ", tresult$p.value, " which is greater than alpha (.05) so we reject H0")
          print(msg)
          print(paste("REJECT NULL HYPOTHESIS: ", ho))
      } else {
          print(paste("FAIL TO REJECT NULL HYPOTHESIS: ", ho))
      }
  }
```

### Load & tidy data. Create 2 groups according to supplement type (orange juice vs ascorbic acid )
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  library(datasets)
  data(ToothGrowth)
  tbl <- data.table(ToothGrowth)
  setnames(tbl, c('len', 'supp', 'dose'), c('tooth_length', 'supplement', 'dose'))
  team_oj <- sqldf("select tooth_length, supplement, dose from tbl where supplement = 'OJ'")
	team_vc <- sqldf("select tooth_length, supplement, dose from tbl where supplement = 'VC'")
	options(digits = 2)
	overview <- rbind(team_oj, team_vc)
	overview$dose <- as.numeric(overview$dose)
```

### explore the data
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  summary(tbl)
  table(tbl$supplement, tbl$dose)
```


### plot to compare tooth growth by supplement type and dose
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  ggplot(overview, aes(x = dose, y = tooth_length)) +
    geom_smooth(data = subset(overview, supplement == "OJ"), aes(colour = "OJ"), linetype = "solid", size = 1) +
    geom_smooth(data = subset(overview, supplement == "VC"), aes(colour = "VC"), linetype = "solid", size = 1) +
    scale_color_manual("Legend", values = c("red", "blue")) +
    labs(title = "Tooth growth by supplement dose") +
    labs(x = "dose", y = "Tooth growth")
```




## Observation: From the 60 samples spread over 2 supplement types with 3 different dosages, Delivery by Orange Juice results in larger tooth growth.  But as dosage increases the growth evens out to where at 2 mg per day the growth between orange juice and ascorbic acid are the same



### output statistics of data
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  stat.desc(team_oj)
  stat.desc(team_vc)
```


## Observation: VC has a higher standard deviation & standard error meaning it is more disbursed.



# Compare tooth growth by supplements and dose


### Does the amount of supplements affect tooth growth?

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  tresult_supplement <- t.test(tooth_length ~ supplement, data = overview, var.equal = FALSE, paired = FALSE, conf.level = .95)
  ho <- 'H0 supplements have not affect on tooth growth'
  ha <- 'Ha supplements will affect tooth growth'
  test_reject(ho, tresult_supplement)
```


### Does the dosage of OJ vs VC affect tooth growth?
##### need to break up the dose into 3 groups for testing
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  dose_small <- subset(overview, dose == 0.5)
  dose_medium <- subset(overview, dose == 1.0)
  dose_high <- subset(overview, dose == 2.0)
```


### small dose
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  tresult_dose_small <- t.test(tooth_length ~ supplement, data = dose_small, var.equal = FALSE, paired = FALSE, conf.level = .95)
  ho <- 'small dose .05 mg is best delivered by a single supplement type to aid tooth growth'
  test_reject(ho, tresult_dose_small)
```


### medium dose
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  tresult_dose_medium <- t.test(tooth_length ~ supplement, data = dose_medium, var.equal = FALSE, paired = FALSE, conf.level = .95)
  ho <- 'medium dose of 1.0 mg is best delivered by a single supplement type to aid tooth growth'
  test_reject(ho, tresult_dose_medium)
```

### high  dose
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
  tresult_dose_high <- t.test(tooth_length ~ supplement, data = dose_high, var.equal = FALSE, paired = FALSE, conf.level = .95)
  ho <- 'high dose of 2.0 mg is best delivered by a single supplement type to aid tooth growth'
  test_reject(ho, tresult_dose_high)
```


# Conclusion: Vitamin C aids in the overall tooth growth of guinea pigs. In smaller doses (.5-1.5) delivering with Orange Juice is more effective  Over larger doses the growth results trend towards similar results per dosage > 2 mg.

# Data may be invalid --> Does the Orange Juice already have Vitamin C in it???
