---
title: "Regression Models Course Project - Is an automatic or manual transmission better for MPG?"
author: "Chris McKelt"
output: pdf_document
---

## Executive Summary

Motor Trend, a popular automotive magazine, has expressed an interest in understanding the relationship between miles per gallon (`mpg`) and transmission type (`am`).  
In order to conduct an analysis in this regard, we will leverage the 1974 Motor Trend Car Road Tests, which can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).
Ultimately, Motor Trend would like to answer two questions:

 1. Is an automatic or manual transmission better for MPG?
 2. Quantify the MPG difference between automatic and manual transmissions?

This analysis was conducted as part of the [Coursera Data Science Specialization](https://www.coursera.org/learn/regression-models/peer/nxntd/regression-models-course-project)

```{r, echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE}
suppressMessages(rm(list = ls()))
suppressMessages(setwd("C:/dev/r-course/6-logistic-regression"))
list.of.packages <- c("tidyverse", "knitr", "markdown", "moments", "data.table", "sqldf", "car")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages, dependencies = TRUE)
sapply(sapply(list.of.packages, library, character.only = TRUE, quietly = TRUE), require, character.only = TRUE, quietly = TRUE)
```

## Data import and tidy

    - vs variable = V engine or a straight engine. 0 means a V-engine, and 1 straight engine. 
    - am variable = automatic / manual 0 = automatic, 1 = manual
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
data(mtcars)
data <- sqldf("select mpg, cyl as cylinders_number, disp as displacement_inches, hp as horsepower, drat as rear_axle_ratio, wt as weight_pounds, qsec as quarter_mile_time, vs as engine_type, am as transmission_type, gear as gears_number, carb as carburetors_number from mtcars")
data <- as_tibble(data)
data$engine_type <- factor(data$engine_type, levels = c(0, 1), labels = c('v_engine', 'straight_engine'))
data$transmission_type <- factor(data$transmission_type, levels = c(0, 1), labels = c('Automatic', 'Manual'))
data$cylinders_number <- factor(data$cylinders_number)
data$carburetors_number <- factor(data$carburetors_number)
```


## Exploratory Analysis

### Conduct a hypothesis test
##### H0 --> transmission_type does not affect miles per gallon
##### H1 --> transmission_type does affect miles per gallon

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
## as variance is not equal Automatic/Manual --> use a Welchs t-test to determine if transmission_type impacts mpg by inspecting the p-value ( alpha = 0.05)
test <- t.test(mpg ~ transmission_type, data = data, var.equal = FALSE, paired = FALSE, conf.level = .95)
test$p.value
```
### 0.001373638  --> since the p-value is less than alpha (0.05) we reject the null hypothesis and can determine there is a cause / effect between transmission_type and mile per gallon



```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
boxplot(mpg ~ transmission_type, data = data,
        col = c("dark green", "light green"),
        xlab = "Transmission",
        ylab = "Miles per Gallon",
        main = "MPG by Transmission Type")
```

## Regression Analysis

We can begin by pulling together a simple linear model directly in line with the first question posed; so, when we look at the relationship between `mpg` and `am` alone, we see 7.24 increase in miles per gallon when comparing just the two variables.

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
model_all <- lm(mpg ~ ., data = data)
vif(model_all)
```
detect collinearity -- find the confounding variables by calculating the variance inflation factors
We can assess the collinearity by variance inflation factor(VIF) 
Values for the VIF that are greater than 40 are considered large. 
We should also pay attention to VIf values over 20
At these point we might consider leaving only one of these variables in the model.
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
vif(model_all)
```

Use AIC to find the best fitting model -->AIC provides a means for model selection.From a set of models the one with minimum AIC value wins
```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
model_best <- step(model_all, direction = "both")
summary(model_best)$coefficients
summary(model_best)$r.squared
```

## Conclusion

## Appendix

Data Table Summary:

```{r}
summary(data)
```

Model Variance Summary:

```{r}
anova(model_all, model_best)
```

Pairs Visual:

```{r message=FALSE, echo=FALSE}
pairs(data, panel = panel.smooth, main = "mtcars data")
```
```{r,echo=FALSE}
kable(data)
```